---
title: "ADS-506 – Final Team Project - Analysis and Forecasting of Retail Sales"
author: "Team 6 - Dipraj Bista, Jessica Hin, Roger Qiu "
date: "Nov 5, 2023"
output: pdf_document
---

# Analysis and Forecasting of Retail Sales

# 1. Intro and EDA

## Data Source

This is a dataset from the U.S. Census Bureau hosted by the Federal Reserve Economic Database (FRED). It is a CSV file that contains sales amounts at 4 different retail stores, each month from 1992 to 2018. The 4 stores are each part of a different sector: furniture, food, used cars and pharmacy. The dataset is available in Kaggle.com here: <https://www.kaggle.com/datasets/census/retail-and-retailers-sales-time-series-collection/data>

## Importing the Data

```{r}

library(fpp2)
library(dplyr)
library(zoo)
library(readr)
library(ggplot2)
library(gridExtra)
# import and inspect the data
df <- read.csv("retail_sales.csv")
head(df)
```

## Time Series Plot

```{r}
# time series plot for each of the 4 stores

# convert to matrix, frequency is 12 months, starts at 1992
furniture_sales <- ts(df$furniture_sales, frequency = 12, start = 1992)
furniture_sales
```

```{r}
plot(furniture_sales, xlab = "Months", ylab = "Furniture sales",  main = "Furniture Sales Over Time")
```

```{r}
food_sales <- ts(df$food_sales, frequency = 12, start = 1992)
plot(food_sales, xlab = "Months", ylab = "Food sales",  main = "Food Sales Over Time")
```

```{r}
used_car_sales <- ts(df$used_car_sales, frequency = 12, start = 1992)
plot(used_car_sales, xlab = "Months", ylab = "Used Car sales",  main = "Used Car Sales Over Time")
```

```{r}
pharmacy_sales <- ts(df$pharmacy_sales, frequency = 12, start = 1992)
autoplot(pharmacy_sales) +
  xlim(2010,2018)
pharmacy_sales
```

## Discussion for EDA

Between the 4 different stores and their area of sales, we can see that there are differences in their seasonality, noise and how they react to certain factors such as the 2008 economic downturn. Overall, sales at all 4 stores have a upward trend. The downturn in 2008 seems to have impacted furniture and food sales the most, followed by used cars. It does not seem to have effected pharmacy sales.

This data can be used to better understand and predict customer purchase behavior and purchase patterns depending on time of the year as well as if there are any correlations between the different sectors. If there are certain months or seasons of the year where a certain store does better, then it would be beneficial to focus any marketing or ads during that period. If there are times were sales have always been slow, then it would be ideal to cut costs during that time such as having a smaller staff or a freeze on marketing during that time.

Given the large time span of this data and the clear visual trends, I believe it can definitely help all 4 of these stores better understand their customers purchasing patterns and behavior and thus help their business.

## Preprocessing

There is not much reprocessing to do for this data set. There are no missing values, the series is not unequally spaced, and no outliers. The only consideration is to use the data only after 2008 because of the housing market crash. Inclusion of years before 2008 can potentially add disruptive information. So let's first re-partition the data for only 2008 and beyond. 

```{r subset the data from 2008 and beyond}
# convert to matrix, frequency is 12 months, starts at 1992
new_df <- df[193:321,]
furn_sales <- ts(new_df$furniture_sales, frequency = 12, start = 2008, end = 2018)
food_sales <- ts(new_df$food_sales, frequency = 12, start = 2008, end = 2018)
car_sales <- ts(new_df$used_car_sales, frequency = 12, start = 2008, end = 2018)
pharm_sales <- ts(new_df$pharmacy_sales, frequency = 12, start = 2008, end = 2018)

# plotting the newly subsetted data
p1 <- autoplot(furn_sales) +
  ggtitle("Furniture Sales Over Time")

p2 <-autoplot(food_sales) +
  ggtitle("Food sales Over Time")

p3 <- autoplot(car_sales) +
  ggtitle("Used Car sales Over Time")

p4 <- autoplot(pharm_sales) +
  ggtitle("Pharmacy sales Over Time")

grid.arrange(p1, p2, p3, p4, nrow = 2, ncol = 2)

```
Now, let's look at the ACF plots first to check for trend and seasonality for each category.
* Furniture sales: Need to twice difference because there's a trend and seasonality component for this category.
* Food Sales: Only need to difference once for the trend component, no obvious seasonality.
* Car sales: Need to twice difference because there's a trend and seasonality component for this category.
* Pharmacy Sales: Need to twice difference because there's a trend and seasonality component for this category.

```{r acf plots}

# acf plots for furniture Sales

acf(furn_sales, lag.max = 120) # there is a positive trend
acf(diff(furn_sales), lag.max = 120) # now you can see there's seasonality @ every lag

# twice differencing the furniture sales series
furn_sales_twice <- diff(diff(furn_sales, lag = 12), lag = 1)

# plotting the new series
p5 <- autoplot(furn_sales_twice) +
  ggtitle("Twice Difference Furniture Sales Over Time")

# acf plots for food sales
acf(food_sales, lag.max = 120) # there is a positive trend
acf(diff(food_sales), lag.max = 120) # no obvious seasonality, just need to difference once

# lag 1 differencing the data
food_sales_once <- diff(food_sales, lag = 1)

p6 <- autoplot(food_sales_once) +
  ggtitle("Differenced Food Sales Over Time")

# acf plots for used car sales
acf(car_sales, lag.max = 120) # there is a positive trend
acf(diff(car_sales), lag.max = 120) # obvious seasonality, just need to difference twice

# lag 1 differencing the data
car_sales_twice <- diff(diff(car_sales, lag = 12), lag = 1)

p7 <- autoplot(car_sales_twice) +
  ggtitle("Twice Differenced Used Car Sales Over Time")

# acf plots for used pharmacy sales
acf(pharm_sales, lag.max = 120) # there is a positive trend
acf(diff(pharm_sales), lag.max = 120) # obvious seasonality, just need to difference twice

# lag 1 differencing the data
pharm_sales_twice <- diff(diff(pharm_sales, lag = 12), lag = 1)

p8 <- autoplot(car_sales_twice) +
  ggtitle("Twice Differenced Pharmacy Sales Over Time")

grid.arrange(p5, p6, p7, p8, nrow = 2, ncol = 2)
```
Finally, we can split the data into a training and test set for each category.

```{r train test split}
# train/test split for furniture sales
furn_train <- window(furn_sales_twice, start = c(2009,2), end = c(2016,12))
furn_test <- window(furn_sales_twice, start = c(2017,1))

# train/test split for food sales
food_train <- window(food_sales_once, start = c(2008,2), end = c(2016,12))
food_test <- window(food_sales_once, start = c(2017,1))

# train/test split for used car sales
car_train <- window(car_sales_twice, start = c(2009,2), end = c(2016,12))
car_test <- window(car_sales_twice, start = c(2017,1))

# train/test split for pharmacy sales
pharm_train <- window(pharm_sales_twice, start = c(2009,2), end = c(2016,12))
pharm_test <- window(pharm_sales_twice, start = c(2017,1))
```


## Modeling

### Naive Forecasts

```{r naive model}

# naive model for furniture sales and plotting it
steps_ahead <- length(furn_test)
naive_furn <- naive(furn_train, h = steps_ahead)
furn_error <- furn_test - naive_furn$mean[steps_ahead]
furn_rmse <- sqrt(mean(furn_error ^2))
autoplot(furn_train) +
  autolayer(naive_furn) +
  ggtitle("Naive Forecasts of Furniture Sales")

# naive model for food sales
naive_food <- naive(food_train, h = steps_ahead)
food_error <- food_test - naive_food$mean[steps_ahead]
food_rmse <- sqrt(mean(food_error^2))
autoplot(food_train) +
  autolayer(naive_food) +
  ggtitle("Naive Forecasts of Food Sales")

# naive model for used car sales
naive_car <- naive(car_train, h = steps_ahead)
car_error <- car_test - naive_car$mean[steps_ahead]
car_rmse <- sqrt(mean(car_error^2))
autoplot(car_train) +
  autolayer(naive_car) +
  ggtitle("Naive Forecasts of Used Car Sales")

# naive model for pharmacy sales
naive_pharm <- naive(pharm_train, h = steps_ahead)
pharm_error <- pharm_test - naive_pharm$mean[steps_ahead]
pharm_rmse <- sqrt(mean(pharm_error^2))
autoplot(pharm_train) +
  autolayer(naive_pharm) +
  ggtitle("Naive Forecasts of Pharmacy Sales")

# table of naive RMSEs

results_tab <- matrix(c(furn_rmse, food_rmse, car_rmse, pharm_rmse), ncol = 1, byrow = TRUE)
rownames(results_tab) <- c('naive_furn','naive_food', 'naive_car', 'naive_pharm')
colnames(results_tab) <- c('RMSE')
results_tab <- as.table(results_tab)
results_tab


```



